Fast Algorithms for Intersection of Non-matching Grids Using \plucker coordinates.

### 2016-09-30 paper_intersections.tex
finished introduction

%\title{Algorithm for Intersections of Nonmatching Grids of Different Dimensions}
\title{Fast Algorithms for Intersection of Nonmatching Grids Using \plucker coordinates.}

@@

%% main text
\section{Introduction}
\notePE{How did you decided for 'non-matching' term? Instead of e.g. incompatible or non-conforming?}

The grid intersection algorithms are crucial for several techniques that try to overcome some limitations of the classical finite element method.
The Chimera method \cite{brezzi_analysis_2001}, also called overset grid, and similar Nitche method \cite{massing_efficient_2013}
allow solution of the problems with changing geometry as in the fluid-structure problems. 
The Mortar method \cite{belgacem_mortar_1999} allows domain decomposition, independent meshing of domains, and supports sliding boundaries.
However our primal motivation is usage of XFEM methods and non-matching meshes of mixed dimension in groundwater models.

The realistic models of groundwater processes including the transport processes and geomechanics have to deal with 
a complex nature of geological formations including the fractures and wells. Although of small scale, these features may have significant impact 
on the global behavior of the system and their representation in the numerical model is imperative.
One possible approach is to model fractures and wells as lower dimensional objects and introduce their coupling with the surrounding continuum. 
The discretization then leads to the meshes of mixed dimensions, i.e. composed of elements of different dimension. This approach 
called mixed-dimensional analysis in the mechanics \cite{bournival_mesh-geometry_2008} is also studied in the groundwater context, see e.g. 
\cite{martin_modeling_2005}, \cite{fumagalli_numerical_2011}, \cite{brezina_analysis_2015} and 
already adopted by some groundwater simulation software, e.g FeFlow \cite{trefry_feflow:_2007} and Flow123d \cite{flow123d}.
Nevertheless as the complexity of the geometry increase (e.g. when lot of fractures are randomly generated) the compatible meshing becomes painful or even 
impossible. In order to avoid these difficulties we may discretize the continuum and every fracture and well independently getting a non-matching 
(or incompatible) mesh of mixed dimensions and then apply XFEM to represent jumps of the solution on the fractures or singularities 
around the wells. The prerequisity for such approach is a fast and robust algorithm for calculating intersections of individual meshes.

We consider a composed mesh $\mathcal T$ consisting of simplicial meshes $\mathcal T_i$ of dimensions $d_i \in \{1,2,3\}$, $i=1,\dots,N_\mathcal{T}$
in the 3d ambient space. We assume that every mesh $\mathcal T_i$ is a connected set with no self intersection.
Further we assume only single 3d mesh $\mathcal{T}_1$.
The mesh intersection problem is to find all pairs of elements $L\in \mathcal{T}_i$, $K\in \mathcal{T}_j$, $i\ne j$ that have non-empty intersection 
and to compute that intersection.

The mesh intersection problem consists of the two parts: First, generate a set of candidate pairs $(K, L)$. Second, compute the intersection for particular pair.
In order to get candidate pairs efficiently, the existing algorithms use either various space trees \cite{massing_efficient_2013} or front tracing
\cite{gander_algorithm_2013} or both \cite{elsheikh_reliable_2012} as in our approach. To compute the actual intersections, we 
use the \plucker coordinates for the line-triangle intersections and a modification of the  Platis and Theoharis algorithm \cite{platis_fast_2003}
for the line-tetrahedron intersections. These are used as the building blocks for the triangle-triangle and the triangle-tetrahedron cases.

\noteJB{Better overview of works, seems that most of works deals with simple intersections, can not find any example of 2d-3d.}

\notePE{Unify the term 'front tracing' 'front tracking' 'advancing front'.}
Our contribution is twofold:  First, we use the front tracking algorithm both to minimize set of intersection candidates as well 
to reuse part of calculations made on neighbouring intersections. Second, we present family of efficient algorithms based on \plucker coordinates 
for computing 1d-2d, 1d-3d, 2d-2d, and 2d-3d intersections of simplicies. The paper is organized as follows \dots

\section{Element Intersections}
\label{sec:elements_intersections}
In this section, we present algorithms for computing intersection of a pair of simplicial elements of a~different dimension in the 3D ambient space. 
In particular we are interested in the intersections of 1D-2D, 1D-3D, 2D-2D, 2D-3D pairs of elements. The fundamental idea is to compute intersection of 1D-2D simplices using the 
\plucker coordinates and reduce all other cases to this one.

@@

@@

\subsection{Tracking boundary of the intersection polygon}

\notePE{It seems to me, that we can really describe the prolongation in general, for 1D and 2D, 
refering to them as components. I tried to do so..}

@@

\section{Advancing Front Method}

add references...
% BIH:
% http://ainc.de/Research/BIH.pdf
% http://diglib.eg.org/handle/10.2312/EGWR.EGSR06.139-149

Consider now a complex mesh of combined dimensions consisting of \emph{components}, which are sets of connected
elements of the same lower dimension (1D, 2D), in the space of connected 3D elements, which we shall call a \emph{bulk}. 
Obtaining all of component-bulk intersections is done in two phases: firstly, we look for the first two elements intersecting each other (initialization); secondly, we prolong the intersection by investigating neighbouring elements (intersection tracking).

To construct the Advancing front algorithm, we shall need:
\begin{itemize}
    \item \textbf{element connectivity} -- we assume this data is available from mesh preprocessing,
    \item \textbf{Axes Aligned Bounding Boxes (AABB)} -- we construct these in order to decide fastly
                    whether to compute the actual intersection of two elements,
    \item \textbf{Bounding Interval Hierarchy (BIH)} -- we alternatively create BIH above AABB to fastly search 
                    created bounding boxes for two coinciding with each other and thus obtaining a candidate pair.
\end{itemize}

The intersection tracking itself can be also seen as a \emph{breadth-first search}
\footnote{Wiki, [online 2016-03-01], \url{https://en.wikipedia.org/wiki/Breadth-first_search}}
algorithm over the BIH, following the component elements.

% \textbf{breadth first search} algorithm:
%   \begin{enumerate}
%     \item\label{en:first} Get next unprocessed component element $k$.
%     \item Find intersection candidates $\mathcal K$ in bulk mesh (3D elements).
%     \item \label{en:q2}For $K\in \mathcal K$ compute intersection $(k, K)$.  
%     \item Push the intersection neigbours into queues: $(k, L) \to Q_b$, $(l, K) \to Q_c$. 
%     \item While $(l, L) \in Q_b$ check intersection $(l, L)$. Append queues.
%     \item Pop pair from $Q_c$. (move to the next component element: go to \ref{en:q2}).
%     \item If $Q_c$ is empty, go to \ref{en:first}.
%   \end{enumerate}

\paragraph{Initialization}
We start with selecting an arbitrary 1D or 2D element. Then we search the bulk elements, checking for 
a collision of bounding boxes, to create a candidate pair. Using only AABB, we need to iterate over
bulk elements in $O(n)$, $n$ being the number of all elements in our case. Using BIH, we can speed 
up the search to $O(\log n)$ on average. In later case, we are paying the costs in the construction of BIH,
which is a quicksort like algorithm running at $O(n\log n)$ on average.

Now that we have provided the first candidate pair, we can look at the scheme in \fig{fig:prolongation} and see us moving from the green box in the left upper corner. If there exists an intersection, we start a new component and
we can procced to tracking the intersection. Otherwise, we select another 1D or 2D element and start over.

%
\begin{figure}[!htb]
%   \vspace{0pt}
  \centering    
    \includegraphics[width=\textwidth]{\figpath prolongation_scheme_ip.pdf}
  \caption{Advancing front algorithm for 1D-2D and 2D-3D intersections.}
  \label{fig:prolongation}
\end{figure}

The advancing front algorithm works the same way for both 1D and 2D components and is displayed by the scheme in \fig{fig:prolongation}. The main idea is to compute intersections for a component element with all 
possible bulk elements, and then move to a~next neighboring component element. For this reason, we define two
queues of candidate pairs: a~\emph{bulk queue} and a~\emph{component queue} (yellow boxes in \fig{fig:prolongation}). We shall now describe the algorithm in detail.

On input we consider a candidate pair, for which a non-empty intersection is computed.
Now we look for new candidate pairs among the neighboring elements (the block Prolongation decide).
Therefore, we iterate over the intersection points and further exploit the topological information.
There are 3 possible cases (applies both for 1D-3D and 2D-3D), how the intersection might be prolonged:
\begin{itemize}
    \item \textbf{IP lies on the component element side and inside the bulk element} \\
            We find all the sides of component element in which the IP lies (IP can be at node and connect more sides).
            Next, we find the component neighboring elements over the sides and push all new candidate pairs
            [component neighbor -- current bulk element] into the component queue.
            Note, that there can be more than one neighbor on a side, if the component has branches.
    \item \textbf{IP lies on the component element side and on the surface of the bulk element} \\
            We find all the faces of bulk element in which the IP lies (1 face, or 2 faces (IP on an edge), 
            or 3 faces (IP at a node)). We find the corresponding neighboring bulk elements over the faces and
            push the new candidate pairs [current component element -- bulk neighbor] into the bulk queue and [component neighbor -- bulk neighbor] into the component queue.
    \item \textbf{IP lies inside component element (therefore must be on the surface of bulk element)} \\
            We proceed as in previous case, but we push only [current component element -- bulk neighbor] candidate pairs, since there is no component neighbor.
\end{itemize}

@@

\section{Benchmarks}

\section{Conclusions}
TODO:
- line intersection tracking for accelerate 2D-2D intersections
- better handling of special cases in paarticular in relation to prolongations
- better calculation reuse (pass with prolongations)
- optimisation of element intersection - skip unnecessary calculations
\section{Acknowledgement}
The paper was supported in part by the Project OP
VaVpI Centre for Nanomaterials, Advanced Technologies  and Innovations
CZ.1.05/2.1.00/01.0005.

>>>>

%\title{Algorithm for Intersections of Nonmatching Grids of Different Dimensions}
\title{Fast Algorithms for Intersection of Non-matching Grids Using \plucker coordinates.}

@@

%% main text
\section{Introduction}
\notePE{How did you decided for 'non-matching' term? Instead of e.g. incompatible or non-conforming?}
\noteJB{Non-matching grid seems to be the most common.}

The grid intersection algorithms are crucial for several techniques that try to overcome some limitations of the classical finite element method.
The Chimera method \cite{brezzi_analysis_2001}, also called overset grid, and similar Nitche method \cite{massing_efficient_2013}
allow solution of the problems with changing geometry as in the fluid-structure problems. 
The Mortar method \cite{belgacem_mortar_1999} allows domain decomposition, independent meshing of domains, and supports sliding boundaries.
However our primal motivation is usage of XFEM methods \cite{fries_extended/generalized_2010} and non-matching meshes of mixed dimension in groundwater models.

The realistic models of groundwater processes including the transport processes and geomechanics have to deal with 
a complex nature of geological formations including the fractures and wells. Although of small scale, these features may have significant impact 
on the global behavior of the system and their representation in the numerical model is imperative.
One possible approach is to model fractures and wells as lower dimensional objects and introduce their coupling with the surrounding continuum. 
The discretization then leads to the meshes of mixed dimensions, i.e. composed of elements of different dimension. This approach 
called mixed-dimensional analysis in the mechanics \cite{bournival_mesh-geometry_2008} is also studied in the groundwater context, see e.g. 
\cite{martin_modeling_2005}, \cite{fumagalli_numerical_2011}, \cite{brezina_analysis_2015} and 
already adopted by some groundwater simulation software, e.g FeFlow \cite{trefry_feflow:_2007} and Flow123d \cite{flow123d}.
Nevertheless as the complexity of the geometry increase (e.g. when lot of fractures are randomly generated) the compatible meshing becomes painful or even 
impossible. In order to avoid these difficulties we may discretize the continuum and every fracture and well independently getting a non-matching 
(or incompatible) mesh of mixed dimensions and then apply XFEM to represent jumps of the solution on the fractures or singularities 
around the wells. The prerequisite for such approach is a fast and robust algorithm for calculating intersections of individual meshes.

We consider a composed mesh $\mathcal T$ consisting of simplicial meshes $\mathcal T_i$ of dimensions $d_i \in \{1,2,3\}$, $i=1,\dots,N_\mathcal{T}$
in the 3d ambient space. We assume that every mesh $\mathcal T_i$ is a connected set with no self intersection.
Further we assume only single 3d mesh $\mathcal{T}_1$.
The mesh intersection problem is to find all pairs of elements $L\in \mathcal{T}_i$, $K\in \mathcal{T}_j$, $i\ne j$ that have non-empty intersection 
and to compute that intersection. The mesh intersection problem consists of the two parts: First, generate a set of candidate pairs $(K, L)$. 
Second, compute the intersection for particular pair.

According to our knowledge there are lot of works using non-matching grids yet only few of them discuss algorithms how to compute their intersections. 
Gander and Japhet \cite{gander_algorithm_2013} present the PANG algorithm for 2d-2d and 3d-3d intersections that can be used e.g. for mesh overlapping methods.
They use the advancing front technique to get candidate pairs in linear time. The algorithm is part of the DUNE library \cite{bastian_towards_2005}.
Massing, Larson, and Logg \cite{massing_efficient_2013} present an algorithm for 2d-3d intersections as part of their implementation of the Nitche method 
which is part of the Doplhin project \cite{Dolphin}. They use axes aligned bounding boxes of elements (AABB) and bounding interval hierarchy (BIH) 
to get intersection candidate pairs of elements, the GTS library \cite{gtslib} is used for 2d-3d intersections. 
Finally, there is the work of Elsheikh and Elsheikh \cite{elsheikh_reliable_2012} presenting an algorithm for 2d-2d mesh union operation which includes 
calculation and imprinting of the intersection curves. They exploit binary space partitioning for search of initial intersection and 
the advancing front method for intersection curve tracking.

In this paper we present a new approach to calculation intersections of simplicial elements of different dimensions based on the \plucker coordinates
further developing the algorithm of Platis and Theoharis \cite{platis_fast_2003} for ray-tetrahedron intersections. Element intersections based on 
\plucker coordinates are combined with the advancing front method which allows us to reuse \plucker coordinates and their products between neigbouring
elements and reduce the number of arithmetic operations. The paper is organized as follows. In Section \ref{sec:element intersecitons} the algorithms for
1d-2d, 1d-3d and 2d-3d intersections of simplices are described. In Section \ref{sec:front advancing} we discuss our implementation of the advancing front 
technique and usage of AABB and BIH for its initialization. Finally, in Section \ref{sec:benchmarks}, we provide benchmarks and comparison of individual algorithms.

\section{Element Intersections}
\label{sec:element intersecitons}
In this section, we present algorithms for computing intersection of a pair of simplicial elements of a~different dimension in the 3D ambient space. 
In particular we are interested in the intersections of 1D-2D, 1D-3D, 2D-2D, 2D-3D pairs of elements. 
The fundamental idea is to compute intersection of 1D-2D simplices using the 
\plucker coordinates and reduce all other cases to this one.

@@

\subsection{Intersection Triangle-Tetrahedron (2D-2D)}
\noteJB{Just note that this algorithm can also be implemented using 1d-2d, but quite complex to not yet fully tested}

@@

\subsection{Tracking boundary of the intersection polygon}
\label{sec:front advancing}
\notePE{It seems to me, that we can really describe the prolongation in general, for 1D and 2D, 
refering to them as components. I tried to do so..}

@@

\section{Front tracking algorithm}
Prerequisites:
  \begin{itemize}
   \item need information about element connectivity (mesh preprocessing)
%   \item mark and check already inspected pairs. Marking elements + hashing.
   \item find the starting intersection -- AABB (Axes Aligned Bounding Boxes) 
   \item possibly create Bounding Interval Hierarchy (BIH)
%   \item Pass already known information with pairs.
%   \item Deal with degenerated cases.
  \end{itemize}

Consider now a complex mesh of combined dimensions consisting of \emph{components}, which are sets of connected
elements of the same lower dimension (1D, 2D), in the space of 3D elements, which we shall call a \emph{bulk}. 
Obtaining all of the 1D-3D and 2D-3D intersections is based on finding the first two elements intersecting each other. Then we can prolongate the intersection by investigating neighbouring elements.

It can be seen as a \emph{breadth-first search}
\footnote{Wiki, [online 2016-03-01], \url{https://en.wikipedia.org/wiki/Breadth-first_search}}
algorithm on the graph of component elements.

\textbf{breadth first search} algorithm:
  \begin{enumerate}
    \item\label{en:first} Get next unprocessed component element $k$.
    \item Find intersection candidates $\mathcal K$ in bulk mesh (3D elements).
    \item \label{en:q2}For $K\in \mathcal K$ compute intersection $(k, K)$.  
    \item Push the intersection neigbours into queues: $(k, L) \to Q_b$, $(l, K) \to Q_c$. 
    \item While $(l, L) \in Q_b$ check intersection $(l, L)$. Append queues.
    \item Pop pair from $Q_c$. (move to the next component element: go to \ref{en:q2}).
    \item If $Q_c$ is empty, go to \ref{en:first}.
  \end{enumerate}

%
\begin{figure}[!htb]
%   \vspace{0pt}
  \centering    
    \includegraphics[width=\textwidth]{\figpath prolongation_scheme_ip.pdf}
  \caption{Prolongation algorithm for 1D-2D and 2D-3D intersections. }
  \label{fig:prolongation}
\end{figure}

The advancing front algorithm works the same way for both 1D and 2D components and is displayed in \fig{fig:prolongation}. The main idea is to compute intersections for a component element with all 
possible bulk elements, and then move to a~next neighboring component element. For this reason, we define two
queues of candidate pairs: a~\emph{bulk queue} and a~\emph{component queue} (yellow boxes in \fig{fig:prolongation}). We shall now describe the algorithm in detail.

On input we consider a candidate pair, for which a non-empty intersection is computed.
Now we look for new candidate pairs among the neighboring elements (the block Prolongation decide).
Therefore, we iterate over the intersection points and further exploit their topological information.
There are 3 possible cases (applies both for 1D-3D and 2D-3D):
\begin{itemize}
    \item \textbf{IP lies on the component element side and inside the bulk element} \\
            We find all the sides of component element in which the IP lies (IP can be at node and connect more sides).
            Next, we find the component neighboring elements over the sides and push all new candidate pairs
            [component neighbor -- current bulk element] into the component queue.
            Note, that there can be more than one neighbor on a side, if the component has branches.
    \item \textbf{IP lies on the component element side and on the surface of the bulk element} \\
            We find all the faces of bulk element in which the IP lies (1 face, or 2 faces (IP on an edge), 
            or 3 faces (IP at a node)). We find the corresponding neighboring bulk elements over the faces and
            push the following new candidate pairs into the bulk queue: [current component element -- bulk neighbor], [component neighbor -- bulk neighbor].
    \item \textbf{IP lies inside component element (therefore must be on the surface of bulk element)} \\
            We proceed as in previous case, but we push only [current component element -- bulk neighbor] candidate pairs, since there is no component neighbor.
\end{itemize}

@@

\section{Benchmarks}
\label{sec:benchmarks}

\section{Conclusions}
\label{sec:conclusins}

TODO:
- line intersection tracking for accelerate 2D-2D intersections
- better handling of special cases in paarticular in relation to prolongations
- better calculation reuse (pass with prolongations)
- optimisation of element intersection - skip unnecessary calculations
%\section{Acknowledgement}
%The paper was supported in part by the Project OP
%VaVpI Centre for Nanomaterials, Advanced Technologies  and Innovations
%CZ.1.05/2.1.00/01.0005.
